---
title: "AMPLIFY masked ML experiment"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

This experiment trains the native R `torch` surrogate and evaluates its predictions under three observation patterns: both `X` and scenario, `X` only, and scenario only. Mean absolute error (MAE) is the primary predictive-performance statistic.

## Production simulation data

The model is trained on the completed outcomes from all 10,000 production simulations per organism scenario. Each simulation contains 1,000 agents; only infected agents enter the individual reproduction-number dataset, including infected agents with zero secondary infections.

```{r}
library(torch)
torch_set_num_threads(min(8L, parallel::detectCores()))
source("../config/simulation.R")
source("../R/build_model.R")
source("../R/extract_outcomes.R")
source("../R/simulate.R")
source("../R/masking.R")
source("../R/torch_model.R")
source("../R/fit_models.R")

config <- default_simulation_config()
study <- load_simulation_batches(
  manifest_path = "../data/derived/manifest.rds",
  root_dir       = ".."
)
run_counts <- table(study$runs$scenario)
if (!identical(as.integer(run_counts[c("lower", "higher")]), c(10000L, 10000L))) {
  stop("The production manifest must contain 10,000 runs per scenario.")
}
agents <- study$agents[study$agents$outcome_complete, ]
rm(study)
invisible(gc())
```

## Run-level train, validation, and test split

Splitting by `run_id` prevents agents from the same simulated epidemic from appearing in multiple partitions. Within each scenario, 70% of runs are assigned to training, 15% to validation, and 15% to final testing.

```{r}
fit <- fit_masked_model(
  agents,
  train_fraction      = 0.70,
  validation_fraction = 0.15,
  hidden_dim_1        = 16L,
  hidden_dim_2        = 8L,
  learning_rate       = 0.01,
  batch_size          = 131072L,
  max_epochs          = 30L,
  patience            = 5L,
  seed                = 20260818L
)

split_summary <- do.call(rbind, lapply(
  c("train", "validation", "test"),
  function(partition) {
    data <- fit$split[[partition]]
    data.frame(
      partition = partition,
      runs = length(unique(data$run_id)),
      infected_agents = nrow(data),
      lower_agents = sum(data$scenario == "lower"),
      higher_agents = sum(data$scenario == "higher")
    )
  }
))
rownames(split_summary) <- NULL
knitr::kable(split_summary)
```

Each training observation is presented under all three masks. The four network inputs are standardized `X`, encoded scenario, and the two observation indicators. The network has hidden widths 16 and 8 and a softplus output trained with Poisson negative log-likelihood.

## Fitting and validation history

```{r}
fit$best_epoch
knitr::kable(head(fit$history, 5), digits = 4)
knitr::kable(tail(fit$history, 5), digits = 4)
```

```{r}
#| fig-cap: "Training and validation Poisson loss by epoch."
#| fig-width: 7
#| fig-height: 4
matplot(
  fit$history$epoch,
  fit$history[c("training_loss", "validation_loss")],
  type = "l",
  lty = 1,
  lwd = 2,
  col = c("#3182bd", "#e6550d"),
  xlab = "Epoch",
  ylab = "Mean Poisson negative log-likelihood"
)
legend(
  "topright",
  legend = c("Training", "Validation"),
  col = c("#3182bd", "#e6550d"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

## Test-set MAE by available inputs

```{r}
mae <- evaluate_masked_mae(fit)
knitr::kable(mae, digits = 3)
```

The overall comparison requested for the prediction interface is:

```{r}
overall_mae <- mae[mae$scenario == "overall", c("pattern", "n", "mae")]
rownames(overall_mae) <- NULL
knitr::kable(overall_mae, digits = 3)
```

Differences between the three MAE values quantify the predictive cost of masking each input. In a completed finite epidemic, the average offspring count across all infected agents is structurally pulled toward one, and unmodeled epidemic timing and susceptible depletion account for substantial individual variation. The network is also fitted with Poisson loss rather than directly minimizing MAE. A later extension could add infection time or susceptible fraction, or define a separate early-phase prediction target, while retaining the same masking interface.

The incomplete-input estimates are population averages over the missing feature. In particular, an `X`-only prediction cannot identify the organism and a scenario-only prediction averages over the training distribution of `X`.

## Save local prediction artifacts

The following generated files are ignored by Git. They allow `README.qmd` to execute its three prediction examples after this report has been rendered.

```{r}
save_masked_model(
  fit,
  weights_path  = "../artifacts/masked_model.pt",
  metadata_path = "../artifacts/masked_model_metadata.rds"
)
```
