---
title: "AMPLIFY masked ML experiment"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

This experiment trains the native R `torch` surrogate with modality dropout and evaluates its predictions under three observation patterns: both `X` and scenario, `X` only, and scenario only. Root mean squared error (RMSE) is the primary predictive-performance statistic.

## Production simulation data

The model uses completed individual outcomes from all 1,000 production simulations per organism scenario. Each simulation contains 10,000 agents observed for 60 days. Recovered infected agents enter the individual reproduction-number dataset, including agents with zero secondary infections; agents still exposed or infectious at the horizon are censored.

```{r}
library(torch)
torch_set_num_threads(min(8L, parallel::detectCores()))
source("../config/simulation.R")
source("../R/build_model.R")
source("../R/extract_outcomes.R")
source("../R/simulate.R")
source("../R/masking.R")
source("../R/torch_model.R")
source("../R/fit_models.R")

config <- default_simulation_config()
study <- load_simulation_batches(
  manifest_path = "../data/derived/production/manifest.rds",
  root_dir       = ".."
)
run_counts <- table(study$runs$scenario)
if (!identical(as.integer(run_counts[c("lower", "higher")]), c(1000L, 1000L))) {
  stop("The production manifest must contain 1,000 runs per scenario.")
}
agents <- study$agents[study$agents$outcome_complete, ]
rm(study)
invisible(gc())
```

## Run-level train, validation, and test split

Splitting by `run_id` prevents agents from the same simulated epidemic from appearing in multiple partitions. Within each scenario, 70% of runs are assigned to training, 15% to validation, and 15% to final testing.

```{r}
fit <- fit_masked_model(
  agents,
  train_fraction      = 0.70,
  validation_fraction = 0.15,
  hidden_dim_1        = 16L,
  hidden_dim_2        = 8L,
  learning_rate       = 0.01,
  batch_size          = 131072L,
  max_epochs          = 30L,
  patience            = 5L,
  seed                = 20260818L
)

split_summary <- do.call(rbind, lapply(
  c("train", "validation", "test"),
  function(partition) {
    data <- fit$split[[partition]]
    data.frame(
      partition = partition,
      runs = length(unique(data$run_id)),
      infected_agents = nrow(data),
      lower_agents = sum(data$scenario == "lower"),
      higher_agents = sum(data$scenario == "higher")
    )
  }
))
rownames(split_summary) <- NULL
knitr::kable(split_summary)
```

During training, torch-native modality dropout independently assigns each observation one of three equally likely states: both inputs retained, scenario dropped, or `X` dropped. Both modalities are never dropped together. The four network inputs are standardized `X`, encoded scenario, and the two observation indicators. Dropout is disabled during validation and inference, where the requested observation pattern is applied explicitly. The network has hidden widths 16 and 8 and a softplus output trained with Poisson negative log-likelihood.

## Fitting and validation history

```{r}
fit$best_epoch
knitr::kable(head(fit$history, 5), digits = 4)
knitr::kable(tail(fit$history, 5), digits = 4)
```

```{r}
#| fig-cap: "Training and validation Poisson loss by epoch."
#| fig-width: 7
#| fig-height: 4
matplot(
  fit$history$epoch,
  fit$history[c("training_loss", "validation_loss")],
  type = "l",
  lty = 1,
  lwd = 2,
  col = c("#3182bd", "#e6550d"),
  xlab = "Epoch",
  ylab = "Mean Poisson negative log-likelihood"
)
legend(
  "topright",
  legend = c("Training", "Validation"),
  col = c("#3182bd", "#e6550d"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

## Test-set RMSE by available inputs

```{r}
rmse <- evaluate_masked_rmse(fit)
knitr::kable(rmse, digits = 3)
```

The overall comparison requested for the prediction interface is:

```{r}
overall_rmse <- rmse[rmse$scenario == "overall", c("pattern", "n", "rmse")]
rownames(overall_rmse) <- NULL
knitr::kable(overall_rmse, digits = 3)
```

Differences between the three RMSE values quantify the predictive cost of masking each input. RMSE is aligned with conditional-mean prediction and therefore with the expected-count interpretation of the Poisson model. Unmodeled epidemic timing and susceptible depletion still account for substantial individual variation. A later extension could add infection time or susceptible fraction while retaining the same modality-dropout interface.

The incomplete-input estimates are population averages over the missing feature. In particular, an `X`-only prediction cannot identify the organism and a scenario-only prediction averages over the training distribution of `X`.

## Save local prediction artifacts

The following generated files are ignored by Git. They allow `README.qmd` to execute its three prediction examples after this report has been rendered.

```{r}
save_masked_model(
  fit,
  weights_path  = "../artifacts/masked_model.pt",
  metadata_path = "../artifacts/masked_model_metadata.rds"
)
```
