---
title: "AMPLIFY multiscale pARG experiment"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

This synthetic experiment connects longitudinal omic observations to a
four-background within-host ODE, explicit reference and effective reproduction
numbers, a summary-coupled carriage ABM, and a missing-modality emulator. It
models pARG carriage and acquisition, not clinical infection.

## Generated study

```{r}
source("../config/multiscale.R")
source("../R/within_host.R")
source("../R/omics.R")
source("../R/reproduction.R")
source("../R/multiscale.R")

study <- readRDS("../data/derived/multiscale/fitted_study.rds")
truth_targets <- combine_multiscale_targets(study$truth)
fitted_targets <- combine_multiscale_targets(study$fitted)
design_summary <- aggregate(
  profile_id ~ site_id,
  study$design,
  length
)
names(design_summary)[2] <- "profiles"
knitr::kable(design_summary, row.names = FALSE)
```

Each of four sites changes the distribution of the three fitted parameters,
contact rate, susceptible fraction, and structural assay availability. Every
host is observed on days 0, 3, 7, 14, and 30.

```{r}
knitr::kable(study$simulated[[1]]$observations, digits = 3, row.names = FALSE)
```

## Truth, observations, and parameter fitting

The ODE uses two *E. coli* and two *Klebsiella* genomic backgrounds. All four
backgrounds share conjugation `h`, division-associated segregation `gamma`, and
antibiotic selection `delta`; a fixed matrix supplies relative within- and
cross-species conjugation rates.

```{r}
parameter_recovery <- do.call(rbind, Map(
  function(truth, fitted) {
    data.frame(
      profile_id = truth$profile$profile_id,
      site_id = truth$profile$site_id,
      parameter = names(truth$parameters),
      truth = unname(truth$parameters),
      fitted = unname(fitted$parameters),
      converged = fitted$fit$convergence == 0,
      boundary = fitted$fit$at_boundary
    )
  },
  study$truth,
  study$fitted
))
parameter_recovery$absolute_error <- abs(
  parameter_recovery$fitted - parameter_recovery$truth
)
parameter_summary <- do.call(rbind, lapply(
  split(parameter_recovery, parameter_recovery$parameter),
  function(rows) data.frame(
    parameter = rows$parameter[1],
    profiles = nrow(rows),
    mean_truth = mean(rows$truth),
    mean_fitted = mean(rows$fitted),
    rmse = sqrt(mean((rows$fitted - rows$truth)^2)),
    boundary_fraction = mean(rows$boundary)
  )
))
knitr::kable(parameter_summary, digits = 3, row.names = FALSE)
```

```{r}
#| fig-cap: "Fitted versus true shared within-host parameters."
#| fig-width: 7
#| fig-height: 4
groups <- split(parameter_recovery, parameter_recovery$parameter)
matplot(
  do.call(cbind, lapply(groups, `[[`, "truth")),
  do.call(cbind, lapply(groups, `[[`, "fitted")),
  pch = 19,
  col = c("#1b9e77", "#d95f02", "#7570b3"),
  xlab = "Truth",
  ylab = "Fitted value"
)
abline(0, 1, lty = 2)
legend(
  "topleft",
  legend = names(groups),
  col = c("#1b9e77", "#d95f02", "#7570b3"),
  pch = 19,
  bty = "n"
)
```

## Within-host reproduction

For donor background `j`, the implementation evolves only the introduced donor
lineage around the pARG-free susceptible trajectory. Recipient acquisitions do
not become donors inside that calculation. The normalized exposure and
first-generation matrix are

$$
\Lambda_{ij}=q^{-1}\int h\omega_{ij}S_i^0(t)R_j^{(1)}(t)/K\,dt,
\qquad
K_{ij}=1-\exp[-p_{est,i}\Lambda_{ij}].
$$

The seed normalization makes the result invariant to the arbitrary numerical
inoculum. Diagonal entries are zero because an already positive background is
not a new genomic background.

```{r}
example <- study$fitted[[1]]$baseline$within
knitr::kable(example$re$lambda, digits = 3, caption = "Normalized transfer exposure")
knitr::kable(example$re$matrix, digits = 3, caption = "Effective within-host next-generation matrix")
data.frame(
  metric = c("R0 within host", "Re within host"),
  value = c(example$r0$reproduction, example$re$reproduction)
) |>
  knitr::kable(digits = 3, row.names = FALSE)
```

## Between-host reproduction

The between-host expectation integrates the complete resistant-fraction
trajectory rather than applying the inoculum function to its mean:

$$
R_{e,i}^{BH}=\int_0^{D_i}c_i s_i p_{est,i}
\{1-\exp[-\kappa p_i(t)]\}\,dt.
$$

The product-form comparator is slightly larger, as expected from the concavity
of the inoculum function.

```{r}
comparison <- fitted_targets[
  fitted_targets$intervention == "baseline",
  c("profile_id", "site_id", "re_between", "product_between")
]
comparison$relative_overstatement <-
  comparison$product_between / comparison$re_between - 1
comparison_summary <- aggregate(
  relative_overstatement ~ site_id,
  comparison,
  function(value) c(mean = mean(value), maximum = max(value))
)
comparison_summary <- data.frame(
  site_id = comparison_summary$site_id,
  mean_overstatement = comparison_summary$relative_overstatement[, "mean"],
  maximum_overstatement = comparison_summary$relative_overstatement[, "maximum"]
)
knitr::kable(comparison_summary, digits = 3, row.names = FALSE)
```

The stochastic ABM retains realized acquisitions as a diagnostic rather than
calling one offspring count a reproduction number.

```{r}
abm_summary <- data.frame(
  infected_carriers = nrow(study$abm$agents),
  realized_acquisitions = sum(study$abm$agents$realized_secondary_acquisitions),
  mean_realized_acquisitions = mean(
    study$abm$agents$realized_secondary_acquisitions
  )
)
knitr::kable(abm_summary, digits = 3, row.names = FALSE)
```

## Paired intervention scenarios

Shorter antibiotic exposure changes the current-condition values but not the
antibiotic-free reference. Conjugation inhibition multiplies `h` by 0.5 and
therefore changes both reference and current-condition values.

```{r}
intervention_summary <- aggregate(
  cbind(r0_within, re_within, r0_between, re_between) ~ intervention,
  fitted_targets,
  mean
)
knitr::kable(intervention_summary, digits = 3, row.names = FALSE)
```

## Missing-modality emulator

The emulator is trained on fitted mechanistic targets from 100 synthetic host
profiles and three intervention settings per profile. During model development,
one model is evaluated on an entirely unseen synthetic site. The browser model
is then retrained with every site represented and evaluated on profiles held out
within each site, because the deployed interface supports all four site
profiles. Errors against fitted targets measure emulation; differences between
fitted and truth targets measure mechanistic fitting error. Neither is a
Bayesian uncertainty interval or a confidence interval.

```{r}
evaluation <- readRDS("../artifacts/multiscale_emulator_evaluation.rds")
complete_input <- evaluation[evaluation$pattern == "all", ]
knitr::kable(complete_input, digits = 3, row.names = FALSE)

profile_missingness <- evaluation[
  evaluation$evaluation_scope == "profile_holdout" &
    evaluation$target %in% c("re_within", "re_between"),
]
knitr::kable(profile_missingness, digits = 3, row.names = FALSE)
```

```{r}
matched <- merge(
  truth_targets,
  fitted_targets,
  by = c("profile_id", "site_id", "intervention"),
  suffixes = c("_truth", "_fitted")
)
targets <- c("r0_within", "re_within", "r0_between", "re_between")
fitting_error <- data.frame(
  target = targets,
  rmse = vapply(targets, function(target) {
    sqrt(mean((
      matched[[paste0(target, "_fitted")]] -
        matched[[paste0(target, "_truth")]]
    )^2))
  }, numeric(1))
)
knitr::kable(fitting_error, digits = 3, row.names = FALSE)
```

## Scientific limitations exposed by the prototype

- The proposal must define and normalize the within-host infectious unit.
- Genomic host switches are interval-censored evidence, not direct observations
  of conjugation.
- Five observations do not identify background-specific biological parameters;
  this prototype therefore fits only three shared values.
- Carriage and clinical infection are different outcomes.
- Deterministic ODEs do not represent rare-transfer extinction.
- Point fitting does not propagate mechanistic uncertainty into the emulator.
- Leave-site-out performance remains weak because the synthetic sites differ in
  both biological parameter distributions and structural modality availability;
  the browser model is a functional demonstration, not a validated predictor
  for an unseen setting.
- The four-versus-five Peru sample schedule, Hi-C detection limit, and transfer
  subscript inconsistency require confirmation before empirical use.
