---
title: "AMPLIFY simulation experiment"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

This report summarizes the full production study for the two antibiotic-resistant organism scenarios. It uses 1,000 independent simulations per scenario with 10,000 agents per simulation and a 120-day horizon. Each simulation draws the agent feature from a standard normal distribution.

## Setup and execution

```{r}
source("../config/simulation.R")
source("../R/build_model.R")
source("../R/extract_outcomes.R")
source("../R/simulate.R")

config <- default_simulation_config()
study <- load_simulation_batches(
  manifest_path = "../data/derived/production/manifest.rds",
  root_dir       = ".."
)
run_counts <- table(study$runs$scenario)
if (!identical(as.integer(run_counts[c("lower", "higher")]), c(1000L, 1000L))) {
  stop("The production manifest must contain 1,000 runs per scenario.")
}
knitr::kable(data.frame(scenario = names(run_counts), runs = as.integer(run_counts)))
```

## Analysis window

A realized individual reproduction number is only informative about the agent who produced it when that agent had a full opportunity to transmit and was infected while the susceptible pool was still essentially intact. Four criteria implement this, and an agent must satisfy all of them to be `analysis_eligible`:

| Criterion | Configuration | Reason |
|---|---|---|
| Infected on or before day `r config$infection_cutoff_day` | `infection_cutoff_day` | Leaves at least `r config$max_days - config$infection_cutoff_day` days to transmit inside the `r config$max_days`-day horizon |
| Transmission window of at least `r config$min_transmission_days` days | `min_transmission_days` | Rules out right-censored offspring counts |
| Infected while at least `r round(100 * config$early_susceptible_fraction)`% of the population was susceptible | `early_susceptible_fraction` | Rules out counts capped by susceptible depletion |
| Not exposed or infectious at the horizon | `max_days` | Rules out agents whose infectious period never ended |

The pseudo-source row identified by `source = -1` is excluded before individual outcomes are constructed. Eligible agents include those with zero secondary infections.

```{r}
eligibility <- do.call(rbind, lapply(
  split(study$agents, study$agents$scenario),
  function(data) {
    data.frame(
      scenario = data$scenario[[1]],
      infected_agents = nrow(data),
      infected_after_cutoff = sum(
        data$infection_day > config$infection_cutoff_day
      ),
      infected_after_early_phase = sum(!data$early_phase),
      censored_at_horizon = sum(!data$outcome_complete),
      eligible_agents = sum(data$analysis_eligible),
      eligible_fraction = mean(data$analysis_eligible)
    )
  }
))
rownames(eligibility) <- NULL
knitr::kable(eligibility, digits = 3)

analysis_agents <- filter_analysis_agents(study$agents)
```

Susceptible depletion, not the infection-day cutoff, is what excludes most `higher`-scenario agents. That scenario leaves its early phase after roughly three weeks and then goes on to infect essentially the whole population well before day `r config$infection_cutoff_day`, so nearly all of its infections happen after susceptible depletion has begun.

```{r}
run_shape <- do.call(rbind, lapply(
  split(study$runs, study$runs$scenario),
  function(data) {
    data.frame(
      scenario = data$scenario[[1]],
      median_early_last_day = stats::median(data$early_last_day),
      median_final_size = stats::median(data$final_epidemic_size),
      runs_active_at_horizon = sum(!data$outcome_complete)
    )
  }
))
rownames(run_shape) <- NULL
knitr::kable(run_shape, digits = 1)
```

## Why the unfiltered average is bounded by one

Every non-seed infection is counted exactly once as somebody's secondary case. Averaging `R_i` over all infected agents therefore returns `(final size - seeds) / final size`, which approaches one from below in any completed epidemic regardless of how transmissible the organism is. The analysis window removes that artifact.

```{r}
comparison <- do.call(rbind, lapply(
  split(study$agents, study$agents$scenario),
  function(data) {
    complete <- data[data$outcome_complete, ]
    within_cutoff <- complete[
      complete$infection_day <= config$infection_cutoff_day,
    ]
    data.frame(
      scenario = data$scenario[[1]],
      all_infected = mean(complete$secondary_cases),
      infected_by_cutoff = mean(within_cutoff$secondary_cases),
      analysis_eligible = mean(
        data$secondary_cases[data$analysis_eligible]
      ),
      target_r0 = config$target_r0[[data$scenario[[1]]]]
    )
  }
))
rownames(comparison) <- NULL
knitr::kable(comparison, digits = 3)
```

## Individual reproduction-number distribution

The target is each eligible agent's realized number of secondary infections, `R_i`, including zero.

```{r}
#' Summarize an individual reproduction-number sample
#'
#' @param data Agent-level simulation data for one scenario.
#'
#' @return A one-row data frame of distribution summaries.
summarize_ri <- function(data) {
  quantiles <- stats::quantile(
    data$secondary_cases,
    probs = c(0.05, 0.25, 0.50, 0.75, 0.95)
  )
  data.frame(
    infected_agents = nrow(data),
    mean_ri = mean(data$secondary_cases),
    sd_ri = stats::sd(data$secondary_cases),
    p05 = quantiles[[1]],
    p25 = quantiles[[2]],
    median = quantiles[[3]],
    p75 = quantiles[[4]],
    p95 = quantiles[[5]],
    zero_fraction = mean(data$secondary_cases == 0)
  )
}

ri_summary <- do.call(
  rbind,
  lapply(split(analysis_agents, analysis_agents$scenario), summarize_ri)
)
ri_summary$scenario <- rownames(ri_summary)
rownames(ri_summary) <- NULL
ri_summary <- ri_summary[c("scenario", setdiff(names(ri_summary), "scenario"))]
knitr::kable(ri_summary, digits = 3)
```

```{r}
#| fig-cap: "Distribution of individual secondary infections by organism scenario."
#| fig-width: 7
#| fig-height: 4
boxplot(
  secondary_cases ~ scenario,
  data = analysis_agents,
  outline = FALSE,
  xlab = "Scenario",
  ylab = "Individual secondary infections (Ri)",
  col = c("#6baed6", "#fd8d3c")
)
```

The next plot retains the large mass at zero and groups the upper tail for readability.

```{r}
#| fig-cap: "Probability mass of individual secondary infections, with values of 7 or more grouped."
#| fig-width: 7
#| fig-height: 4
ri_group <- pmin(analysis_agents$secondary_cases, 7L)
distribution <- prop.table(
  table(analysis_agents$scenario, factor(ri_group, levels = 0:7)),
  margin = 1
)
barplot(
  distribution,
  beside = TRUE,
  names.arg = c(as.character(0:6), "7+"),
  legend.text = rownames(distribution),
  args.legend = list(x = "topright", bty = "n"),
  xlab = "Individual secondary infections (Ri)",
  ylab = "Fraction of infected agents",
  col = c("#fd8d3c", "#6baed6")
)
```

## Achieved early-epidemic reproduction

Calibration targets the mean `R_i` of analysis-eligible agents. The table reports both the pooled eligible-agent mean and the distribution of run-level eligible means.

```{r}
pooled_early <- do.call(rbind, lapply(
  split(analysis_agents, analysis_agents$scenario),
  function(x) {
    data.frame(
      scenario = x$scenario[[1]],
      mean_ri = mean(x$secondary_cases),
      se = stats::sd(x$secondary_cases) / sqrt(nrow(x)),
      infected_agents = nrow(x)
    )
  }
))
run_early <- do.call(rbind, lapply(split(study$runs, study$runs$scenario), function(x) {
  data.frame(
    scenario = x$scenario[[1]],
    mean_run_ri = mean(x$eligible_mean_ri, na.rm = TRUE),
    sd_run_ri = stats::sd(x$eligible_mean_ri, na.rm = TRUE),
    runs = nrow(x)
  )
}))
rownames(pooled_early) <- NULL
rownames(run_early) <- NULL
knitr::kable(pooled_early, digits = 3)
knitr::kable(run_early, digits = 3)
```

The pooled means sit close to the 0.5 and 4.0 targets. The run-level `lower` mean is smaller than the pooled mean because most `lower` runs go extinct with only their five seeds, so each of those runs contributes a mean of zero with equal weight. These values vary around the targets because this is a finite stochastic model. The calibration script refines the scenario intercepts before the production run.
