---
title: "AMPLIFY simulation experiment"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

This report runs a reproducible inspection study for the two antibiotic-resistant organism scenarios. It uses 100 independent simulations per scenario with 1,000 agents per simulation. The production experiment increases this to 10,000 simulations per scenario using the same functions.

## Setup and execution

```{r}
source("../config/simulation.R")
source("../R/build_model.R")
source("../R/extract_outcomes.R")
source("../R/simulate.R")

config <- default_simulation_config()
study <- run_simulation_study(
  config,
  n_reps  = 100L,
  workers = 1L
)
```

All reported individual outcomes have complete follow-up through the fixed 250-day safety horizon.

```{r}
with(study$runs, table(scenario, outcome_complete))
```

## Individual reproduction-number distribution

The target is each infected agent's realized number of secondary infections, `R_i`, including zero.

```{r}
#' Summarize an individual reproduction-number sample
#'
#' @param data Agent-level simulation data for one scenario.
#'
#' @return A one-row data frame of distribution summaries.
summarize_ri <- function(data) {
  quantiles <- stats::quantile(
    data$secondary_cases,
    probs = c(0.05, 0.25, 0.50, 0.75, 0.95)
  )
  data.frame(
    infected_agents = nrow(data),
    mean_ri = mean(data$secondary_cases),
    sd_ri = stats::sd(data$secondary_cases),
    p05 = quantiles[[1]],
    p25 = quantiles[[2]],
    median = quantiles[[3]],
    p75 = quantiles[[4]],
    p95 = quantiles[[5]],
    zero_fraction = mean(data$secondary_cases == 0)
  )
}

ri_summary <- do.call(
  rbind,
  lapply(split(study$agents, study$agents$scenario), summarize_ri)
)
ri_summary$scenario <- rownames(ri_summary)
rownames(ri_summary) <- NULL
ri_summary <- ri_summary[c("scenario", setdiff(names(ri_summary), "scenario"))]
knitr::kable(ri_summary, digits = 3)
```

```{r}
#| fig-cap: "Distribution of individual secondary infections by organism scenario."
#| fig-width: 7
#| fig-height: 4
boxplot(
  secondary_cases ~ scenario,
  data = study$agents,
  outline = FALSE,
  xlab = "Scenario",
  ylab = "Individual secondary infections (Ri)",
  col = c("#6baed6", "#fd8d3c")
)
```

The next plot retains the large mass at zero and groups the upper tail for readability.

```{r}
#| fig-cap: "Probability mass of individual secondary infections, with values of 7 or more grouped."
#| fig-width: 7
#| fig-height: 4
ri_group <- pmin(study$agents$secondary_cases, 7L)
distribution <- prop.table(
  table(study$agents$scenario, factor(ri_group, levels = 0:7)),
  margin = 1
)
barplot(
  distribution,
  beside = TRUE,
  names.arg = c(as.character(0:6), "7+"),
  legend.text = rownames(distribution),
  args.legend = list(x = "topright", bty = "n"),
  xlab = "Individual secondary infections (Ri)",
  ylab = "Fraction of infected agents",
  col = c("#fd8d3c", "#6baed6")
)
```

## Early-epidemic reproduction

Calibration uses agents infected while at least 90% of the population remains susceptible. The table reports both the pooled early-agent mean and the distribution of run-level early means.

```{r}
early_agents <- study$agents[study$agents$early_phase, ]
pooled_early <- do.call(rbind, lapply(split(early_agents, early_agents$scenario), function(x) {
  data.frame(
    scenario = x$scenario[[1]],
    mean_ri = mean(x$secondary_cases),
    se = stats::sd(x$secondary_cases) / sqrt(nrow(x)),
    infected_agents = nrow(x)
  )
}))
run_early <- do.call(rbind, lapply(split(study$runs, study$runs$scenario), function(x) {
  data.frame(
    scenario = x$scenario[[1]],
    mean_run_ri = mean(x$early_mean_ri),
    sd_run_ri = stats::sd(x$early_mean_ri),
    runs = nrow(x)
  )
}))
rownames(pooled_early) <- NULL
rownames(run_early) <- NULL
knitr::kable(pooled_early, digits = 3)
knitr::kable(run_early, digits = 3)
```

These values are expected to vary around the targets because this is a finite stochastic model. The calibration script refines the scenario intercepts with larger Monte Carlo batches before the production run.
